Characterizing Signal Loss, Error, and Bias in the 21cm Reionization \\Power Spectrum: A Revised Study of PAPER-64

### 2017-12-23 PAPER.tex
removed bootstrapping text about baseline sampling

The following techniques each have advantages when it comes to maximizing sensitivity and understanding systematics in 
data. However, some have limitations, and we will discuss circumstances in which there are trade-offs. We choose to focus on 
these four techniques because they represent major steps in PAPER's power spectrum pipeline, with several of them also 
being standard steps in general $21$ cm analyses.
\begin{itemize}
\item \textbf{Fringe-rate filtering:} Fringe-rate filtering is an averaging scheme for time-ordered data 
(\citealt{parsons_et_al2016}). Broadly, a fringe-rate filter increases the sensitivity of a dataset and reduces the number of 
independent samples by an amount dependent on the width of the averaging window. However, it can also affect the presence 
of foregrounds and systematics. We explain the trade-offs of filtering in more detail in Section \ref{sec:toymodel_frf}.
\item \textbf{Weighting:} A dataset can be weighted to emphasize certain features and minimize others. One particular flavor of 
weighting employed by PAPER is inverse covariance weighting in frequency, which is a generalized version of inverse variance 
weighting that also takes into account frequency correlations (\citealt{liu_tegmark2011}; \citealt{dillon_et_al2013a}; \citealt{liu_et_al2014a}; \citealt{liu_et_al2014b}). This weighting has the effect of down-weighting correlated 
information (i.e. foregrounds) and up-weighting noise-like information (i.e. EoR). However, a challenge of inverse covariance 
weighting is in accurately defining a covariance matrix that best describes the data. We investigate the impact of different types of weighting on signal loss in Section \ref{sec:SiglossOverview}.
\item \textbf{Bootstrapping:} In addition to using theoretical models for covariance matrices and theoretical error estimation 
methods, bootstrapping is one way to estimate errors. Bootstrapping is a useful method for estimating errors of a dataset from 
itself (\citealt{andrae2010}). By randomly drawing many samples of the data, we obtain a sense of its inherent variance, though there are subtleties to 
consider such as the independence of values in a dataset. We explore potential pitfalls of bootstrapping in Section \ref{sec:ErrorOverview}.
\item \textbf{Jackknife testing:} A resampling technique useful for estimating bias, jackknives can be taken along different 
dimensions of a dataset to cross-validate results. In particular, null tests can be used to verify whether results are free of 
systematics, as done with CO power spectra (\citealt{keating_et_al2016}) and CMB measurements (see e.g. \citealt{ade_et_al2008}; \citealt{chiang_et_al2010}; \citealt{bischoff_et_al2011}; \citealt{das_et_al2011}; \citealt{araujo_et_al2012}; \citealt{crites_et_al2015}; \citealt{ade_et_al2016}; \citealt{ade_et_al2017}; \citealt{sherwin_et_al2017}). An EoR detection must pass jackknife and null tests, which we highlight in Section \ref{sec:JackknifeOverview}.
\end{itemize}

@@

For PAPER's analysis, we choose to compute error bars on our measurements that have been derived from its inherent 
variance. A common technique used to do this is bootstrapping. For pedagogical purposes, we first define the technique of 
bootstrapping and then illustrate some of its pitfalls through toy models.

@@

We will now discuss a second subtle feature of bootstrapping that fails to maximize the sensitivity of a dataset. Suppose we 
have $5$ independent measurements of the sky (from $5$ different baselines, for example). A bootstrap then consists of $5$ 
measurements that are drawn randomly with replacement from the original set. If all $5$ spaces are filled randomly, there is a 
high probability that some measurements will be repeated in the bootstrap because they are drawn more than once. The 
bootstrap may therefore consist of only $3$ or $4$ independent measurements --- a number smaller than the total number of 
samples. In fact, the probability of drawing $5$ completely independent measurements is less than $4\%$.

@@

The difference between the two methods is most pronounced for datasets with large numbers of elements. As a dataset grows 
in size, it becomes increasingly rare to draw entirely independent samples, and thus the benefit of ensuring, say, $999$ 
independent samples out of $1000$ is most noticeable. The more elements in our dataset, the more sensitivity we can gain by 
mandating that most of our measurements are independent ones. An analog for the y-axis in Figure \ref{fig:toy_error2} is 
therefore the sensitivity of a final power spectrum measurement.

In summary, bootstrapping can be an effective and straightforward way to estimate errors of a dataset. However, we have 
illustrated two situations in which bootstrapping can lead to under-estimated errors and sensitivities. First, we have shown that 
bootstrapped error depends strongly on the number of elements drawn in a bootstrap sample. Estimated errors can drop to 
arbitrarily small values when the number of elements drawn exceeds the effective number of independent elements. We have 
also shown a situation in which a bootstrapping method does not provide the most sensitive measurement possible. While 
bootstrapping is convenient because it provides a way to estimate errors from the data itself, one must assess whether certain 
analysis choices have compromised the method and whether a variation of traditional re-sampling could be preferred instead.

\begin{figure}
	\centering
	\includegraphics[trim={0cm 0cm 0cm 0cm},width=\columnwidth]{plots/toy_error2.png}
	\caption{The fraction of independent elements in one bootstrap sample as a function of the total number of elements in a 
dataset (i.e. if $4$ out of $5$ elements are independent, this would yield a fraction of $0.8$). A fraction of $1.0$ means that all 
elements in the bootstrap sample are independent, and therefore sensitivity is maximized (alternately, this axis can be thought 
of as power spectrum sensitivity, where sensitivity increases moving upwards along the y-axis). Two bootstrapping methods are shown here --- sampling all elements with replacement (gray) and sampling only the last element with replacement (black). The former fails to maximize sensitivity since some elements end up being repeated due to random sampling.} 
%The `outlier' points lying at a fraction of $1.0$ at the far left of the plot occur in the off-chance that the last spot in the sample happens to be the one missing element instead of a repeated element --- this is more likely to happen with a small total number of elements but also results in a more dramatically different fraction as compared to the case of a repeated element in the last spot, hence why these points appear like outliers.}
	\label{fig:toy_error2}
\end{figure}

@@

We bootstrap PAPER-64 power spectra in order to determine confidence intervals for our results. In this section, we highlight 
two major changes in the way we estimate errors since A15, using the lessons we've learned about sampling with replacement 
and bootstrapping independent samples.

@@

Given our new understanding of the sensitivity of bootstraps to the number of elements sampled, we have removed the second 
bootstrapping step along time entirely and now simply bootstrap over baselines with the ``single replacement" strategy 
described in Section \ref{sec:ErrorOverview}.

The single replacement strategy also provides a higher sensitivity estimate as only at most one baseline is repeated in each 
group. In A15, baselines were chosen for groups randomly with replacement. In doing so, $\sim3$$-4$ baselines per group 
were repeated (Figure \ref{fig:toy_error2} shows the fraction of independent samples to be $\sim$$63\%$ for $10$ total 
independent samples) which sacrificed some sensitivity.

Power spectrum estimates with these bootstrapping changes for PAPER-64 fringe-rate filtered data are shown in Figure 
\ref{fig:data_errors}. The estimates are unweighted in order to disentangle the effects of bootstrapping from signal loss. As 
shown in the figure, when the baseline average includes many repeated samples our sensitivity is not maximized (blue), while when 
only one sample in the same average is allowed the possibility of being a repeat, the error on the noise-dominated higher $k$ 
modes converges to the theoretical noise level (black).  However, if more elements are drawn for each bootstrap than the number of 
independent samples (by over-sampling elements along the time axis), repeated values begin to crop up and the apparent 
variation between bootstraps drops, resulting in limits below the predicted noise level (red).

@@

\begin{itemize}
\item Signal loss, previously found to be $<2\%$ in A15, was under-estimated by a factor of $\sim$$1000$ for inverse 
covariance weighting. For our new analysis, we use a regularized covariance weighting method to minimize loss (Section 
\ref{sec:Weight}). However, because our revised weighting method is not as aggressive as the former, our results are still a 
factor of $\sim7$ mK higher than previous limits. Under-estimated signal loss therefore represents the bulk of our revision. We 
note that PAPER's analysis is not the first to under-estimate loss; results from the GMRT (\citealt{paciga_et_al2013}) were also 
revised from new signal loss calculations associated with their singular value decomposition foreground filter.
\item Power spectrum errors, originally computed by bootstrapping, were under-estimated by a factor of $\sim$$2$ due to 
oversampling data whose effective number of independent samples was reduced from fringe-rate filtering. Our limits were also 
over-estimated by a factor of $\sim$$2$ from using a suboptimal way of combining baselines into groups when forming power 
spectra (Section \ref{sec:Boot}). Hence, the net change for our error estimation is negligible, though the rationale behind the 
previous estimation was incorrect.
\item Several factors used in an analytic expression to predict the noise-level in PAPER-64 data were revised, yielding a 
decrease in predicted sensitivity level by a factor of $\sim$$2$ in mK (Section \ref{sec:PSSense}). We have verified our revised 
prediction extensively using pure noise simulations. We note that our sensitivity prediction is revised by a factor less than our 
power spectrum result, implying that if taken at face value, the theoretical prediction for noise in A15 was too high for its data 
points.
\end{itemize}

>>>>

The following techniques each have advantages when it comes to maximizing sensitivity and understanding systematics in 
data. However, some have limitations, and we will discuss circumstances in which there are trade-offs. We choose to focus on 
these four techniques because they represent major steps in PAPER's power spectrum pipeline, with several of them also 
being standard steps in general $21$ cm analyses.
\begin{itemize}
\item \textbf{Fringe-rate filtering:} Fringe-rate filtering is an averaging scheme for time-ordered data 
(\citealt{parsons_et_al2016}). Broadly, a fringe-rate filter increases the sensitivity of a dataset and reduces the number of 
independent samples by an amount dependent on the width of the averaging window. However, it can also affect the presence 
of foregrounds and systematics. We explain the trade-offs of filtering in more detail in Section \ref{sec:toymodel_frf}.
\item \textbf{Weighting:} A dataset can be weighted to emphasize certain features and minimize others. One particular flavor of 
weighting employed by PAPER is inverse covariance weighting in frequency, which is a generalized version of inverse variance 
weighting that also takes into account frequency correlations (\citealt{liu_tegmark2011}; \citealt{dillon_et_al2013a}; \citealt{liu_et_al2014a}; \citealt{liu_et_al2014b}). This weighting has the effect of down-weighting correlated 
information (i.e. foregrounds) and up-weighting noise-like information (i.e. EoR). However, a challenge of inverse covariance 
weighting is in accurately defining a covariance matrix that best describes the data. We investigate the impact of different types of weighting on signal loss in Section \ref{sec:SiglossOverview}.
\item \textbf{Bootstrapping:} In addition to using theoretical models for covariance matrices and theoretical error estimation 
methods, bootstrapping is one way to estimate errors. Bootstrapping is a useful method for estimating errors of a dataset from 
itself (\citealt{andrae2010}). By randomly drawing many samples of the data, we obtain a sense of its inherent variance, though there are subtleties to 
consider such as the independence of values in a dataset. We explore this potential pitfall of bootstrapping in Section \ref{sec:ErrorOverview}.
\item \textbf{Jackknife testing:} A resampling technique useful for estimating bias, jackknives can be taken along different 
dimensions of a dataset to cross-validate results. In particular, null tests can be used to verify whether results are free of 
systematics, as done with CO power spectra (\citealt{keating_et_al2016}) and CMB measurements (see e.g. \citealt{ade_et_al2008}; \citealt{chiang_et_al2010}; \citealt{bischoff_et_al2011}; \citealt{das_et_al2011}; \citealt{araujo_et_al2012}; \citealt{crites_et_al2015}; \citealt{ade_et_al2016}; \citealt{ade_et_al2017}; \citealt{sherwin_et_al2017}). An EoR detection must pass jackknife and null tests, which we highlight in Section \ref{sec:JackknifeOverview}.
\end{itemize}

@@

For PAPER's analysis, we choose to compute error bars on our measurements that have been derived from its inherent 
variance. A common technique used to do this is bootstrapping. For pedagogical purposes, we first define the technique of 
bootstrapping and then illustrate one of its pitfalls through a toy model.

@@

\begin{comment} % REMOVED THIS HALF OF BOOTSTRAPPING SECTION
We will now discuss a second subtle feature of bootstrapping that fails to maximize the sensitivity of a dataset. Suppose we 
have $5$ independent measurements of the sky (from $5$ different baselines, for example). A bootstrap then consists of $5$ 
measurements that are drawn randomly with replacement from the original set. If all $5$ spaces are filled randomly, there is a 
high probability that some measurements will be repeated in the bootstrap because they are drawn more than once. The 
bootstrap may therefore consist of only $3$ or $4$ independent measurements --- a number smaller than the total number of 
samples. In fact, the probability of drawing $5$ completely independent measurements is less than $4\%$.

@@

The difference between the two methods is most pronounced for datasets with large numbers of elements. As a dataset grows 
in size, it becomes increasingly rare to draw entirely independent samples, and thus the benefit of ensuring, say, $999$ 
independent samples out of $1000$ is most noticeable. The more elements in our dataset, the more sensitivity we can gain by 
mandating that most of our measurements are independent ones. An analog for the y-axis in Figure \ref{fig:toy_error2} is 
therefore the sensitivity of a final power spectrum measurement.
\end{comment}

In summary, bootstrapping can be an effective and straightforward way to estimate errors of a dataset. However, we have 
illustrated a situation in which bootstrapping can lead to under-estimated errors and therefore under-estimated power spectrum limits. We have shown that 
bootstrapped error depends strongly on the number of elements drawn in a bootstrap sample. Estimated errors can drop to 
arbitrarily small values when the number of elements drawn exceeds the effective number of independent elements. 
%We have also shown a situation in which a bootstrapping method does not provide the most sensitive measurement possible. 
While bootstrapping is convenient because it provides a way to estimate errors from the data itself, one must assess whether certain 
analysis choices have compromised the method and whether a variation or an avoidance of traditional re-sampling could be preferred instead.

%\begin{figure}
%	\centering
%	\includegraphics[trim={0cm 0cm 0cm 0cm},width=\columnwidth]{plots/toy_error2.png}
%	\caption{The fraction of independent elements in one bootstrap sample as a function of the total number of elements in a 
%dataset (i.e. if $4$ out of $5$ elements are independent, this would yield a fraction of $0.8$). A fraction of $1.0$ means that all 
%elements in the bootstrap sample are independent, and therefore sensitivity is maximized (alternately, this axis can be thought 
%of as power spectrum sensitivity, where sensitivity increases moving upwards along the y-axis). Two bootstrapping methods are shown here --- sampling all elements with replacement (gray) and sampling only the last element with replacement (black). The former fails to maximize sensitivity since some elements end up being repeated due to random sampling.} 
%The `outlier' points lying at a fraction of $1.0$ at the far left of the plot occur in the off-chance that the last spot in the sample happens to be the one missing element instead of a repeated element --- this is more likely to happen with a small total number of elements but also results in a more dramatically different fraction as compared to the case of a repeated element in the last spot, hence why these points appear like outliers.}
%	\label{fig:toy_error2}
%\end{figure}

@@

We bootstrap PAPER-64 power spectra in order to determine confidence intervals for our results. In this section, we highlight 
one major changes in the way we estimate errors since A15, using the lesson we've learned about bootstrapping independent samples.

@@

Given our new understanding of the sensitivity of bootstraps to the number of elements sampled, we have removed the second 
bootstrapping step along time entirely and now simply bootstrap over baselines. Power spectrum estimates with this bootstrapping change for PAPER-64 fringe-rate filtered data are shown in Figure 
\ref{fig:data_errors}. The estimates are unweighted in order to disentangle the effects of bootstrapping from signal loss. As 
shown in the figure, when more elements are drawn for each bootstrap than the number of 
independent samples (by over-sampling elements along the time axis), repeated values begin to crop up and the apparent variation between bootstraps drops, resulting in limits below the predicted noise level (gray).

@@

\begin{itemize}
\item Signal loss, previously found to be $<2\%$ in A15, was under-estimated by a factor of $\sim$$1000$ for inverse 
covariance weighting. For our new analysis, we use a regularized covariance weighting method to minimize loss (Section 
\ref{sec:Weight}). However, because our revised weighting method is not as aggressive as the former, our results are still a 
factor of $\sim7$ mK higher than previous limits. Under-estimated signal loss therefore represents the bulk of our revision. We 
note that PAPER's analysis is not the first to under-estimate loss; results from the GMRT (\citealt{paciga_et_al2013}) were also 
revised from new signal loss calculations associated with their singular value decomposition foreground filter.
\item Power spectrum errors, originally computed by bootstrapping, were under-estimated by a factor of $\sim$$1.4$ in mK due to oversampling data whose effective number of independent samples was reduced from fringe-rate filtering (Section \ref{sec:Boot}). 
\item Several factors used in an analytic expression to predict the noise-level in PAPER-64 data were revised, yielding a 
decrease in predicted sensitivity level by a factor of $\sim$$2$ in mK (Section \ref{sec:PSSense}). We have verified our revised 
prediction extensively using pure noise simulations. We note that our sensitivity prediction is revised by a factor less than our 
power spectrum result, implying that if taken at face value, the theoretical prediction for noise in A15 was too high for its data 
points.
\end{itemize}
